---
title: "Citibike"
author: "Tarun Bhadri, Ben Boeke, Taeyoung Justin Yun, Victor Manuel Jauregui-Tapia, Vincent Qu Dee"
date: "11/30/2021"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```
```{r, cache=TRUE}
library(tidyverse)
library(gganimate)
library(geosphere)
library(lubridate)
library(ggplot2)
library(dplyr)
library(knitr)
library(leaflet)
```


## Data frame setup 
### Data Cleaning 
```{r}
#Start New RMD File for Analysis
#Reading the new csv file 
citisample <- read.csv("citibike.csv")
str(citisample)
citisample$X <- NULL
citisample$start.station.id <- as.factor(citisample$start.station.id)
citisample$end.station.id <- as.factor(citisample$end.station.id)
citisample$bikeid <- as.factor(citisample$bikeid)
citisample$usertype <- as.factor(citisample$usertype)
citisample$gender <- as.factor(citisample$gender)
# Fix gender
citisample$gender <- ifelse(citisample$gender == 1, "male", ifelse(citisample$gender == 2, "female", "unknown"))
citisample$gender <- as.factor(citisample$gender)
# Create a column for approximate age
citisample$age <- 2021 - citisample$birth.year
```

### Data Cleaning Continued
```{r}
citisample$starttime <- as.POSIXct(strptime(citisample$starttime, "%Y-%m-%d %H:%M:%S"))
citisample$stoptime <- as.POSIXct(strptime(citisample$stoptime, "%Y-%m-%d %H:%M:%S"))

citisample$starthour <- hour(citisample$starttime)
citisample$day <- date(citisample$starttime)

citisample$month <- as.factor(month(citisample$day))

citisample$numWeekday <- wday(citisample$starttime)
citisample$dayid <- as.factor(ifelse(citisample$numWeekday < 6, "Weekday", "Weekend"))
citisample$weekNum <- week(citisample$starttime)
citisample$distmeters <- distHaversine(cbind(citisample$start.station.latitude, citisample$start.station.longitude), cbind(citisample$end.station.latitude, citisample$end.station.longitude))

#rename weekdays
citisample$numWeekday <- ifelse(citisample$numWeekday==1,"Sunday",citisample$numWeekday)
citisample$numWeekday <- ifelse(citisample$numWeekday==2,"Monday",citisample$numWeekday)
citisample$numWeekday <- ifelse(citisample$numWeekday==3,"Tuesday",citisample$numWeekday)
citisample$numWeekday <- ifelse(citisample$numWeekday==4,"Wednesday",citisample$numWeekday)
citisample$numWeekday <- ifelse(citisample$numWeekday==5,"Thursday",citisample$numWeekday)
citisample$numWeekday <- ifelse(citisample$numWeekday==6,"Friday",citisample$numWeekday)
citisample$numWeekday <- ifelse(citisample$numWeekday==7,"Saturday",citisample$numWeekday)
citisample$numWeekday <- factor(citisample$numWeekday,levels = c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"))

citisample$speed <- citisample$distmeters / citisample$tripduration
str(citisample)
```
### Merging NYC Weather Data
```{r}
#Merging Datasets
weather <- read.csv("NYCWeather2019.csv")
weather$STATION <- NULL
weather$NAME <- NULL
weather$DATE <- as.Date(weather$DATE, format= "%m/%d/%Y")
weather$TAVG <- (weather$TMAX + weather$TMIN)/2

citiday <- citisample %>%
  group_by(day, gender, dayid) %>%
  summarize(count = n(),
            dist = mean(distmeters, na.rm = TRUE),
            dur = mean(tripduration, na.rm = TRUE),
            speed = dist/dur)

combined_gender <- merge(citiday, weather, by.x = "day", by.y = "DATE")
```

## Overview 

The goal of this report is to increase citi bike usage across the city. To accomplish this goal we must understand trends in the data and how it effects bike usage. To start we have provided some context on the citibike data. Next, we will analyze the impact of asymmetric traffic on bike station usage. We will dive deeper into bike usage to understand maintenance schedules. Finally, we analyze the impact of weather on bike usage. With these insights we will be able to provide citibike with recommendations to increase usage across the city. 


## Citi Bike Data Context 
```{r}
ggplot(citisample, aes(x=month)) + geom_bar() +labs( title = "Amount of Rides in 2019 per month")
```

```{r}
ggplot(citisample, aes(x=month, fill = gender)) + geom_bar(position=position_dodge()) +labs( title = "Amount of Rides in 2019 by gender")
```

```{r}
ggplot(citisample, aes(x=usertype)) + geom_bar(position=position_dodge()) +labs( title = "Subscriber v. Customer amount of rides in 2019")
```

```{r}
citisample %>%
  group_by(numWeekday) %>%
  summarize(count = n(),
            dist = mean(distmeters, na.rm = TRUE),
            dur = mean(tripduration, na.rm = TRUE),
            speed = dist/dur) %>%
  ggplot(aes(x=numWeekday, y = dur/60, fill = count)) + geom_col() +labs(y="Ride duration in Minutes", x = "", title = "Avg trip duration each day" )
```





## Assymetric Traffic 
```{r}
citisample$route <- paste(citisample$start.station.name,citisample$end.station.name,sep=" -> ")

citisample %>% 
  group_by(route) %>% 
  summarize(count=n()) %>% 
  arrange(desc(count)) %>% 
  top_n(n=5) %>% kable(format="markdown")
```

This table shows the top 5 most popular routes from our sample 2019 data. Management should consider these routes when allocating bikes across stations. One should expect the ending station to have a surplus of bikes compared to the departing station. Our team will continue to analyze which stations actually have a surplus or deficit. 
```{r}
bike_departures <- group_by(citisample, station = `start.station.name`, latitude = `start.station.latitude`, longitude = `start.station.longitude`)

departure <- summarise(bike_departures, departure_count = n())


bike_arrivals <- group_by(citisample, station = `end.station.name`, latitude = `end.station.latitude`, longitude = `end.station.longitude`)
arrival <- summarise(bike_arrivals, arrival_count = n())

#merge departure and arrival data into one df
bike_deficit <- merge(departure, arrival, all = TRUE)
# make NA's = 0
bike_deficit[is.na(bike_deficit)] <- 0

#make deficit variable: == arrivals - departures
bike_deficit$deficit <- bike_deficit$departure_count - bike_deficit$arrival_count
```

```{r}
#Now we are just going to show the 10% of most problematic stations. That means that we are only going to show the 10% of stations where the difference between arrivals and departures is higher
bike_deficit10percent<-sort(abs(bike_deficit$deficit),decreasing = TRUE)
minimum<-bike_deficit10percent[length(bike_deficit10percent)%/%10]

bike_deficit_worststations<-bike_deficit[abs(bike_deficit$deficit)>=minimum,]



# circle size representative of the magnitude of the deficit or surplus the station has (absolute val of how many more departures it has than arrivals)
# Red circles have a net deficit, blue circles have a net surplus
# click on station to see station name and net deficit or surplus
leaflet(bike_deficit_worststations) %>% 
  addTiles() %>%
  addCircleMarkers(lng = bike_deficit_worststations$longitude, lat = bike_deficit_worststations$latitude, 
                   popup = paste(bike_deficit_worststations$station, "<br>", ifelse(bike_deficit_worststations$deficit>=0, "Bike deficit = ", "Bike surplus = "), 
                                abs(bike_deficit_worststations$deficit)), 
                   radius = abs(bike_deficit_worststations$deficit)/5, color = ifelse(bike_deficit_worststations$deficit>0, "red", "green"))
```
This map shows the top 10% problematic stations. 

```{r}
bike_deficit_10<-arrange(bike_deficit, (deficit))[1:10,]


def_icon <- makeIcon("red_bike.png", iconWidth = 24, iconHeight =32)

leaflet(bike_deficit_10) %>% 
  addTiles() %>%
  addMarkers(lng = ~longitude, lat = ~latitude, label=paste(bike_deficit_10$station,bike_deficit_10$deficit), icon = def_icon)


surp_icon <- makeIcon("green_bike.png", iconWidth = 24, iconHeight =32)
bike_surplus_10<-arrange(bike_deficit, -deficit)[1:10,]


leaflet(bike_surplus_10) %>% 
  addTiles() %>%
  addMarkers(lng = ~longitude, lat = ~latitude, label=paste(bike_surplus_10$station,bike_surplus_10$deficit), icon = surp_icon)

```

This map marks the top 10 stations with either a surplus or deficit. On each marker you can also see the amount for that station. The station of interest would be North Moore St & Greenwich st which has the largest deficit at -123. The station with the highest surplus would E 13 st and Ave A with  147 bikes. 
Looking at the maps shows that the top stations with deficits are in the midlower half of Manhattan. The top station with surplus are shift to the right in the north east part of Manhattan. Management should consider understanding the moment of bikes and transport bikes back to the western side where there is a deficit. 


## Bike Usage 
### Bike ID
### Top 10 Most Used Bikes
```{r}
topbikes <- citisample %>%
  group_by(bikeid) %>%
  summarize(count=n()) %>%
  arrange(desc(count)) %>%
  top_n(n=10) %>% kable(format="markdown")
topbikes
```

### How Usertype Affects Ridership
```{r}
#time of day by usertype
citisample %>% ggplot(aes(x=starthour,group=usertype,fill=usertype)) + geom_density(alpha=.3) + theme(legend.title=element_blank(),axis.text.y=element_blank())+ggtitle(expression(atop("How Usertype Affects Time of Day Distribution",atop("Customer = 24 hour/7 Day pass  | Subscriber = Annual Pass")))) 
```

Our data shows the different ride patterns between customers and subscribers. Subscribers were more likely to use bikes during rush hour times/used the mode of transportation for work purposes. Regular customers' usage of bikes steadily increased until a peak at 15:00. Perhaps this can be explained by users riding a bike for lunch break or post-lunch slump.

### Hourly Distribution of Rides
```{r}
citisample %>% ggplot(aes(x=starthour,fill=factor(numWeekday))) + geom_density(alpha=.2)+facet_wrap(~numWeekday,ncol=2)+theme(legend.position="none",axis.text.y=element_blank(),plot.title=element_text(hjust=.5)) + ggtitle(expression(atop("Hourly Distribution of Rides",atop("Weekday Peaks during Morning Rush (8am-9am) and Afternoon Rush (5pm-6pm)"))))
```
We looked at hourly distribution of rides since we believe that it is valuable information for Citibike in the cases they need live data on the conditions of bikes and when they can expect a lot of usage by both subscribers and customers. We observed weekday peeks during Morning Rush (8-9am) and Afternoon Rush (5-6pm)

## Weather 
### How does the average temperature affect the distance rode?

```{r}
ggplot(combined_gender, aes(x = TAVG, y = dist)) + geom_point(size = 2, alpha = 0.5, color = "yellowgreen") + geom_smooth(color = "darkgreen") + ylim(0,3000) 
```
It seems until about 75 degrees Fahrenheit, distance rode increases, meaning people prefer to ride the CitiBike when it is not swelteringly warm. This could be because they do not want to be sweaty in their nicer clothes for work.


### How does the amount of precipitation affect the distance rode?
```{r}
str(combined_gender)
combined_gender %>%
  group_by(PRCP) %>%
  summarize(count = n(),
            dist = mean(dist, na.rm = TRUE),
            avgprcp = mean(PRCP, na.rm = TRUE),
            speed = dist/dur) %>%
  ggplot(aes(x = PRCP,y = dist)) + geom_point() + geom_smooth()

```

As precipitation increases there is a weak reduction in distance rode on a Citibike. This makes sense as most users of Citibikes are commuters and one-day users are likely more sensitive to the amount of rain as they are more likely to be tourists.

### Relationship between average temperature and distance
```{r}
combined_gender %>%
  group_by(TAVG) %>%
  summarize(count = n(),
            dist = mean(dist, na.rm = TRUE),
            avgprcp = mean(PRCP, na.rm = TRUE),
            speed = dist/dur) %>%
  ggplot(aes(x = TAVG,y = dist)) + geom_point() + geom_smooth()
```

This graph shows the average distance ridden for each recorded temperature. The ridership tends to ride a longer distance in higher temperatures, with a pretty clear positive correlation shown by the trend line. Citi gets a larger ridership in the summer, so they need to stock more bikes at different stations to reduce individual bike usage and maintainence costs.


### Does the ridership distance differ between months? 
```{r}
citisample %>%
  group_by(month) %>%
  summarize(count = n(),
            dist = mean(distmeters, na.rm = TRUE),
            dur = mean(tripduration, na.rm = TRUE),
            speed = dist/dur) %>%
  ggplot(aes(x=month, y = dist, fill = count)) + geom_col() +labs(y="Distance", title = "Avg Distance each Month" )
```

This graph shows how the average distance of riders differs by month. This is similar to the previous graph, in that riders tend to ride for a longer distance in the summer months, which are warmer, and shorter in the colder months. Again, this creates the issue that Citibike needs to consider reducing bike supply in the winter months and increasing bike supply in the summer months, as more distance creates more wear and tear so more bikes would be optimal to ensure there are enough bikes to satisfy demand. 




## Conclusion and Key actionable Insights & Recommendations

### Recommendation 1

The idea of our business solution here is to even out the distribution of bikes across NYC, and hopefully take some bikes from stations with surplus and give them to the stations with deficits. Additionally, in order to assist with the peaks of when customers are using CitiBikes, specifically rush hours at 9 AM and 5 PM when the workday begins and ends, customers riding bikes right before those times, at 8 AM and 4 PM per se, would be offered credit to end their rides at stations that usually face deficits during rush hour peaks. Ideally, this would help CitiBike better manage the peaks when they face the most business as well as the most issues. 

### Recommendation 2

Since subscribers are our main target audience, providing exclusive benefits for these individuals will only further promote Citibike’s services. One example of this is creating a dashboard on its app that allows subscribers to see data about rush hour in real time and providing them with an option to reserve a bike at the nearest station up to 30 minutes before the scheduled time. Since we identified that subscribers who use the services use it as a mode of transportation to and from their jobs, we believe this exclusive benefit will encourage other commuters to look into biking as a more healthy and viable alternative.